When to trust phylogeographic inferences?

### 2024-03-24 manuscript/phylogeo.tex
edits

\documentclass[aps,rmp, onecolumn]{revtex4}
%\documentclass[a4paper,10pt]{scrartcl}
%\documentclass[aps,rmp,twocolumn]{revtex4}

@@

In a typically Bayesian phylogeographic inference problem as discussed in \citet{pybus_unifying_2012}, the input data are typically viral genome sequences as well as their sampling times and sampling locations.
Software like BEAST will then sample the posterior distribution of the phylogenetic tree topology, the sampling times and locations of ancestral nodes in the tree, and model parameters like the diffusion constant $D$ and the evolutionary rate $\mu$.
Results can be estimates of the ancestral location of a particular groups of samples, or summary statistics of the dispersal process.
Popular summaries include empirical estimates of the diffusion constant  \citep{pybus_unifying_2012} and a so-called \emph{dispersal velocity} \citep{dellicour_using_2017}. These are calculated from the estimated displacements along branch $i$ with parent $p_i$ given by $d_i = |\rvec_{p_i} - \rvec_{i}|$ and the elapsed time $\Delta t_i$ as
\begin{equation}
    D_w = \frac{\sum_{i\in B}d_i^2}{4\sum_{i\in B} \Delta t_i} \quad \mathrm{and}  \quad v_w = \frac{\sum_{i\in B} d_i}{\sum_{i\in B} \Delta t_i}
\end{equation}
where $B$ is the set of all branches of the tree.
Effectively $D_w$ divides the sum of all observed squared displacements by the total time elapsed on the tree to obtain and estimate of $D$.
The dispersal velocity estimate compares the sum of absolute values of observed displacements to the total time, which has dimensions of a velocity.
Alternative formulations use the mean of fractions $d_i^2/4t_i$ or $d_i/t_i$ instead of the ratio of sums, but these tend to be less robust.

The purpose of this note is to explore how sensitive phylogeographic inferences are to model assumptions and what quantities can be robustly estimated.
My focus here is only on the ability and robustness to estimate dispersal parameter and ancestral locations and I will assume perfect knowledge of the tree and the sampling locations.
So no tree-sampling will be necessary.

Fig.~\ref{fig:D_and_v} shows empirical estimates of $D_w$ and $v_w$ as defined above from simulated data using freely diffusing locations along Kingman or Yule trees.
The diffusion constant $D$ can be estimated robustly from these data, but $v_w$ is not a well-defined quantity.
Values of $v_w$ depends strongly on the sample size and there is no ground truth to compare to as there is no parameter with dimensions of a velocity in the model.
That $v_w$ is ill-defined is not surprising: By changing the sample size, the average length of branches changes. Diffusion along each branch results in a displacement of $d_i \sim \sqrt{4Dt_i}$ such that a quantity that depends on $\Delta d_i / t_i \sim 1/t_i^{1/2}$ will change with the time $t_i$ that elapsed along the branch.
If the underlying dispersal process is diffusive, estimates of $v_w$ are not meaningful.

\begin{figure}[tb]
    \includegraphics*[width=0.48\textwidth]{figures/yule_total_dispersal.png}
    \includegraphics*[width=0.48\textwidth]{figures/kingman_total_dispersal.png}
    \caption{\label{fig:D_and_v}{\bf Empirical summary statistics of $D$ and ``dispersal velocity'' for different sample sizes.}
    While diffusion constant $D$ (equal to 1 in this example) can be robustly estimated from simulation data of the underlying model, the dispersal rate $v_w$ is not a well-defined quantity. Its numerical value depends strongly on sample size and there is no ground truth to compare to. This sample size dependence is present for Yule trees (left) and Kingman trees (right). }
\end{figure}

\section*{How does density regulation affect estimates phylogeographic estimates?}
Phylogeographic inference typically assumes that the spatial diffusion process is independent of the branching pattern of the tree and ignores coupling between the two that might arise from density dependent growth rates.
If one simulates a diffusive spatial process in a square of length $L$, the population density shows strong fluctuations whenever the diffusion constant $D<L^2/T_c$, see Fig.~\ref{fig:heterogeneity} where $T_c$ is the coalescent times scale of the population.
In a well-mixed neutral model, $T_c$ is equal to the population size $N$.

@@

\begin{figure}[tb]
    \includegraphics*[width=\textwidth]{figures/heterogeneity_free_diffusion.pdf}
    \caption[short]{\label{fig:heterogeneity}{\bf Population density variation in free diffusion models.}
    When spatial motion is independent of population density and coalesce is sufficiently rapid, the realized population density fluctuates strongly across the habitat. The left panel shows the population density for the case $D = 0.1 L^2/N$ in linear bins of size $L/20$. The right panels shows the standard deviation of the population density for different bin sizes as a function of the diffusion coefficient. For $D>L^2/N$, the heterogeneity approaches the expected value in the well mixed limit. }
\end{figure}

Biological populations spread into areas that support them and will quickly populate fertile areas that are below their carrying capacity.
The simplest model of this process are FKPP waves that model the population density $\phi(x,t)$ as \citep{fisher_wave_1937,KPP1937,hallatschek_life_2010}
\begin{equation}
    \frac{\partial \phi(x,t)}{\partial t} = D\frac{\partial^2 \phi(x,t)}{\partial x^2} + \alpha \phi(x,t)(1-\phi(x,t)/\phi_0)
\end{equation}
where $\phi_0$ is the carrying capacity and $\alpha$ is the population growth rate at zero density.
This equation admits solutions with a traveling front, that the population ``invades" the empty territory with velocity $v = 2\sqrt{D \alpha}$.
Note that the model itself does not have an intrinsic velocity.
The speed at which the population invades space is given as compound of diffusion and accelerated growth in regions of low density and only grows like the square root of diffusion constant.
It is thus not clear a priori how estimates of $D$ will be affected by density regulation, even in a constant habitat.

Dependence of growth rate (and thus tree structure) on density is one key element that phylogeographic models ignore because it is very difficult to capture mathematically.
It is, however, fairly straightforward to include this in simulations and these simulations can be used to investigate the nature of phylogeographic inferences when some of the assumptions are violated.
We calculate the population density using a Gaussian kernel with interaction radius $r$
\begin{equation}
    \phi(\rvec) = \sum_{i\in N} \frac{e^{-|\rvec_i - \rvec|^2/2r^2}}{N\sqrt{2\pi r^2}}
\end{equation}
where the sum is over all individuals $i$ in the population of size $N$.
As soon as the interaction radius $r$ is much smaller than the habitat size, this density regulation leads to a more homogeneous population density, see Fig.~\ref{fig:density_reg}.
Diffusion constants estimated from the population remain accurate with moderate overestimation of small diffusion constants.
But with decreasing diffusion, the population fragments into smaller than smaller subpopulations that don't interact or coalesce.
This manifests in an increasing time to the MRCA, see Fig.~\ref{fig:density_reg}.
Spatial location and the coalescence process are thus strongly coupled, in conflict with model assumptions and estimates of population processes are likely biased.

\begin{figure}
    \includegraphics[width=0.48\textwidth]{figures/density_reg_heterogeneity.pdf}
    \includegraphics[width=0.48\textwidth]{figures/density_reg_tmrca.pdf}
    \caption{\label{fig:density_reg} {\bf Density regulation reduces fluctuations in density regulation, but results in population fragmentation at low diffusion constants.}}
\end{figure}

>>>>

\documentclass[aps,rmp, twocolumn]{revtex4}
%\documentclass[a4paper,10pt]{scrartcl}
%\documentclass[aps,rmp,twocolumn]{revtex4}

@@

In a typically Bayesian phylogeographic inference problem as discussed in \citet{pybus_unifying_2012}, the input data are typically viral genome sequences as well as their sampling times and sampling locations.
Software like BEAST will then sample the posterior distribution of the phylogenetic tree topology, the sampling times and locations of ancestral nodes in the tree, and model parameters like the diffusion constant $D$ and the evolutionary rate $\mu$.
Results can be estimates of the ancestral location of a particular groups of samples, or summary statistics of the dispersal process.
Popular summaries include empirical estimates of the diffusion constant  \citep{pybus_unifying_2012} and a so-called \emph{dispersal velocity} \citep{dellicour_using_2017}. These are calculated from the estimated displacements along branch $i$ with parent $p_i$ given by $d_i = |\rvec_{p_i} - \rvec_{i}|$ and the elapsed time $\Delta t_i$ as
\begin{equation}
    D_w = \frac{\sum_{i\in B}d_i^2}{4\sum_{i\in B} \Delta t_i} \quad \mathrm{and}  \quad v_w = \frac{\sum_{i\in B} d_i}{\sum_{i\in B} \Delta t_i}
\end{equation}
where $B$ is the set of all branches of the tree.
Effectively $D_w$ divides the sum of all observed squared displacements by the total time elapsed on the tree to obtain and estimate of $D$.
It is known as the `weighted' estimate of the diffusion constant.
The weighted dispersal velocity estimate compares the sum of absolute values of observed displacements to the total time, which has dimensions of a velocity.
Alternative formulations use the mean of fractions $D_b = |B|^{-1} \sum_{i\in B}d_i^2/4t_i$ or $v_b = |B|^{-1} \sum_{i\in B}d_i/t_i$ instead of the ratio of sums. \
We refer to these here as `by branch' estimates. 
The latter tend to be dominated by short branches and thus noisier.

The purpose of this note is to explore how sensitive phylogeographic inferences are to model assumptions and what quantities can be robustly estimated.
My focus here is only on the ability and robustness to estimate dispersal parameter and ancestral locations and I will assume perfect knowledge of the tree and the sampling locations.
So no tree-sampling will be necessary.
To explore qualitatively different behaviors, I will consider Kingman and Yule tree ensembles. Kingman trees are the classical neutral model of a constant size population, while Yule trees capture features of growing populations (see Fig.~\ref{fig:illustration_tree}).

\begin{figure*}[tb]
    \includegraphics*[width=0.7\textwidth]{figures/illustration_tree.pdf}
    \caption{\label{fig:illustration_tree}{\bf Illustration Yule and Kingman trees and the spatial location of the tips.}
    The total tree length of Yule trees is proportional to the number of tips $n$ and the time to the MRCA is $\log(n)$.
    Kingman trees have a total tree length that is proportional to $\log(n)$ and are characterized by rapid merging close to the present. }
\end{figure*}

Fig.~\ref{fig:D_and_v} shows empirical estimates of diffusion constants and velocity as defined above from simulated data using freely diffusing locations along Kingman trees.
The diffusion constant $D$ can be estimated robustly from these data and are compatible with the diffusion constant used to simulate the data $D=1$.
Values of $v_w$ and $v_b$, however, depend strongly on the sample size. 
That the velocity estimates are ill-defined is not surprising: By changing the sample size, the average length of branches changes. Diffusion along each branch results in a displacement of $d_i \sim \sqrt{4Dt_i}$ such that a quantity that depends on $\Delta d_i / t_i \sim 1/t_i^{1/2}$ will change with the time $t_i$ that elapsed along the branch.
With more samples in the tree, the average branch length decreases and the velocity estimates increase. 
This expected scaling $\sim \sqrt{n}$ is indicated by the dashed line in Fig.~\ref{fig:D_and_v}.
For Yule trees, the average branch length is independent of the number of samples and the velocity estimates are more stable. 
But estimates for $v_w$ and $v_b$ are not consistent with each other and their numerical values have no interpretation within the model.

In either case, estimates of $v_w$ or $v_b$ are not meaningful. 
The underlying model is diffusive and has no parameters that individually or combined would have dimensions of a velocity. 
There is no ground truth to compare the estimates to and $v_w$ and $v_b$ are just descriptive summaries of spatial spread with unclear interpretation.

\begin{figure}[tb]
    \includegraphics*[width=0.48\textwidth]{figures/kingman_dispersal.pdf}
    \caption{\label{fig:D_and_v}{\bf Empirical summary statistics of $D$ and ``dispersal velocity'' for different sample sizes.}
    While diffusion constant $D$ (equal to 1 in this example) can be robustly estimated from simulation data of the underlying model, the dispersal rate $v_w$ (weighted) and $v_b$ (by branch) are not well-defined quantities. Their numerical value depends strongly on sample size and there is no ground truth to compare to. The expected scaling $\sim \sqrt{n}$ of $v$ estimates is indicated by the dashed line.}
\end{figure}

\section*{How does density regulation affect estimates phylogeographic estimates?}
Phylogeographic inference typically assumes that the spatial diffusion process is independent of the branching pattern of the tree and ignores coupling between the two. 
Such a coupling could, for example, arise when growth rates depend on local population density, and it has been known for a long time that ignoring such coupling leads to very unrealistic population densities \citep{felsenstein_pain_1975}.
If one simulates a diffusive spatial process in a square of length $L$ with periodic boundary conditions, the population density shows strong fluctuations whenever the diffusion constant $D<L^2/T_c$, where $T_c$ is the coalescent times scale of the population (see Fig.~\ref{fig:heterogeneity}).
In a well-mixed neutral model, $T_c$ is equal to the population size $N$.

@@

\begin{figure*}[tb]
    \includegraphics*[width=\textwidth]{figures/heterogeneity_free_diffusion.pdf}
    \caption[short]{\label{fig:heterogeneity}{\bf Population density variation in free diffusion models.}
    When spatial motion is independent of population density and coalesce is sufficiently rapid, the realized population density fluctuates strongly across the habitat. The left panel shows the population density for the case $D = 0.1 L^2/N$ in linear bins of size $L/20$. The right panels shows the standard deviation of the population density for different bin sizes as a function of the diffusion coefficient. For $D>L^2/N$, the heterogeneity approaches the expected value in the well mixed limit. }
\end{figure*}

Biological populations spread into areas that support them and will quickly populate fertile areas that are below their carrying capacity.
A simple model for the population density $\phi(x,t)$ would be 
\begin{equation}
    \label{eq:FKPP}
    \frac{\partial \phi(x,t)}{\partial t} = D\frac{\partial^2 \phi(x,t)}{\partial x^2} + \alpha \phi(x,t)(1-\phi(x,t)/\phi_0)
\end{equation}
where $\phi_0$ is the carrying capacity and $\alpha$ is the population growth rate at zero density.
This model has a stable fixed point at $\phi=\phi_0$.
In a static environment, we thus expect the population density to settle at a constant density $\phi_0$ and density fluctuations will decay with rate $\alpha$.

While dependence of growth (and thus tree structure) on density is a key component of real-world population dynamics, it is difficult to implemented in phylogeographic inference frameworks.
It is, however, fairly straightforward to include density dependence in forward simulations and these simulations can be used to investigate the effect of such coupling between growth and spatial location on phylogeographic inferences.
We calculate the population density using a Gaussian kernel with interaction radius $r$
\begin{equation}
    \phi(\rvec) = \sum_{i\in N} \frac{e^{-|\rvec_i - \rvec|^2/2r^2}}{N\sqrt{2\pi r^2}}
\end{equation}
where the sum is over all individuals $i$ in the population of size $N$.

As soon as the interaction radius $r$ is much smaller than the habitat size, this density regulation leads to a more homogeneous population density, see Fig.~\ref{fig:density_reg}.
Diffusion constants estimated from the population remain accurate with moderate overestimation of small diffusion constants.
But with decreasing diffusion, the population fragments into smaller than smaller subpopulations that don't interact or coalesce.
This manifests in an increasing time to the MRCA, see Fig.~\ref{fig:density_reg}.
Spatial location and the coalescence process are thus strongly coupled, though this coupling manifests itself in population structure rather than estimates of the diffusion coefficients.

\begin{figure*}
    \includegraphics[width=0.48\textwidth]{figures/density_reg_heterogeneity.pdf}
    \includegraphics[width=0.48\textwidth]{figures/density_reg_tmrca.pdf}
    \caption{\label{fig:density_reg} {\bf Density regulation reduces fluctuations in density regulation, but results in population fragmentation at low diffusion constants.} The left panel shows population heterogeneity measured using bins of linear dimension $L/5$ for $\alpha =0.05$ for different values of the interaction radius $r$. For $r\ll L$, density fluctuations are strongly suppressed. At the same time, the time to the most recent common ancestor increases dramatically, indicating that the population fragments in subpopulations that don't interact.}
\end{figure*}

The simplest model of this process are FKPP waves that model the population density $\phi(x,t)$ as \citep{fisher_wave_1937,KPP1937,hallatschek_life_2010}

This equation admits solutions with a traveling front, that the population ``invades'' the empty territory with velocity $v = 2\sqrt{D \alpha}$.
Note that the model itself does not have an intrinsic velocity.
The speed at which the population invades space is given as compound of diffusion and accelerated growth in regions of low density and only grows like the square root of diffusion constant.
It is thus not clear a priori how estimates of $D$ will be affected by density regulation, even in a constant habitat.
